Does the Price Stability of 'No Name' Mean No Price Increases? A Strategy by Loblaws to Retain Customers

### 2024-11-26 paper/paper.qmd
intro, abstract, title completed

---
title: "Title"
subtitle: "Subtitle"
author: 
  - Chenming Zhao
thanks: "Code and data are available at: [https://github.com/ChenmingZ2000/Loblaws-NoNameBrand-Grocery-Price-Trend-Analysis.git](https://github.com/ChenmingZ2000/Loblaws-NoNameBrand-Grocery-Price-Trend-Analysis.git).
        We gratefully acknowledge Jacob Filipp’s groceries dataset used in this analysis. The data is available at: [https://jacobfilipp.com/hammer/](https://jacobfilipp.com/hammer/)"
date: today
date-format: long
abstract: "abstract"
format: pdf
number-sections: true
bibliography: references.bib
#header-includes:
#  - \usepackage{tabularray/tabularray}
pdf-engine: lualatex
---

@@

@@

To simulate, test, analyze, and clean the polling data, the statistical programming language R was employed [@citeR]. Key libraries that supported the data analysis include `tidyverse` [@tidyverse], `dplyr` [@dplyr], `arrow` [@arrow], `knitr` [@knitr], `ggplot2` [@ggplot2], `kableExtra` [@kableExtra] and `patchwork` [@patch]. Additionally, the `rstanarm` [@rstanarm] library was utilized to implement multilevel modeling for prediction purposes. Also this analysis inherits the folder structure from Professor Rohan Alexander from University of Toronto [@rohan].

## Measurement
In the real world, pricing data for consumer goods is a dynamic reflection of market trends, regional preferences, and competitive strategies. The prices of products fluctuate based on supply chain disruptions, seasonal demand, and retail strategies, making it essential to track and analyze this information systematically. The Hammer dataset serves as a resource in understanding these trends, offering a detailed collection of product pricing data sourced from leading retailers like Loblaws, Walmart, and Metro. This dataset is embedded with complexities arising from the data collection process, competitive pricing strategies, and evolving consumer behavior.

Although the data collection process for [@hammer] data is not explicitly documented, based on the emphasis on "related" content on the Project Hammer website and standard industry methods, it is highly likely that common techniques such as web scraping, API integration, and data standardization were employed. These are typical approaches used to construct structured datasets from online retail platforms. That is, retailers provide public access to their product inventories and prices through their websites, making them prime candidates for automated data extraction. For example, web scraping tools are deployed to extract product details such as names, brands, vendors, categories, unit prices, and quantities, while timestamps capture the exact moment of data retrieval to ensure time-sensitive accuracy. However, even with automated methods, the process is not without its challenges. Retailers often introduce barriers like CAPTCHA systems or dynamic content loading, which require robust solutions such as headless browsers or API integrations to ensure smooth data acquisition.

Sampling and representativeness are considerations in Hammer's data collection process. Since the dataset aims to reflect national pricing trends, it includes a diverse range of vendors and product types to capture regional and competitive variations. However, inherent biases may still exist. For instance, not all vendors update their online product catalogs in real time, which may introduce lag in capturing price changes. Additionally, regional pricing disparities might not always align with the dataset's collection schedule, potentially skewing the data toward more actively monitored regions or vendors.

Once the raw data is collected, it undergoes cleaning and standardization. Product names, for instance, often include variations due to marketing strategies (e.g., "Club Size" or "Family Pack"), which are standardized to ensure consistency across entries. Similarly, data with missing values—such as prices or unit information—are flagged for further investigation.

@@

@fig-2 shows how average prices vary over time, segmented by product `type` (left) and `size` (right). In the left panel, we observe distinct pricing patterns across product types. For instance, "Seasoning" demonstrates higher volatility compared to "Flour/Rice," likely due to differences in demand elasticity. The "Milk" category shows occasional price dips, possibly influenced by the introduction of new products or promotional events.

The right panel shows price trends segmented by size. "Extra Large" products generally have higher average prices but exhibit more unstable trends over time. In contrast, "Small" products show minimal price variation, possibly due to their inherently lower pricing or because their unit price is relatively higher compared to larger sizes.

@@

```{r, message=FALSE, warning=FALSE, fig.width= 7, fig.height= 3}
#| label: fig-2
#| fig-cap: Average Support Rate  for Donald Trump from 2024-07-21 to 2024-10-23 reported by different pollsters
#| echo: false
# Group data by date and product type, and calculate average price for each type per date
type_trend_data <- loblaws_data %>%
  group_by(date, type) %>%
  summarise(avg_price = mean(current_price, na.rm = TRUE)) %>%
  ungroup()

@@

The goal of our modeling strategy is to build a pricing model aimed at predicting changes in `current_price` over time while considering key variables that influence price fluctuations. The structure of the model combines fixed effects and random effects and includes multiple variables, such as `product_id`, product `size`, and product `type`. This modeling approach is designed to balance the complexity of the data while ensuring robust predictions. The hierarchical structure of the model allows us to capture variations at the level of individual products while identifying overall trends across different product sizes and types. It also enables us to forecast future price trends of Loblaws "No Name" groceries, observing whether the retailer's pricing behavior aligns with their commitments.

Here, we provide a brief overview of the Bayesian analytical model used to investigate the price trends of Loblaws "No Name" products. Background details, model assumptions, and diagnostics are included in [Appendix -@sec-model-details].

@@

$u_{\text{productID}[i]}$ is the random intercept for each product ID, capturing product-specific variations in pricing.

@@

The model was implemented using the rstanarm package and employed MCMC sampling for Bayesian inference. To validate the model, we conducted out-of-sample testing and calculated the predictive accuracy using RMSE (Root Mean Square Error), which was 0.83 on the log-transformed price scale. In the context of grocery price data, this level of error is acceptable, as market forces and external factors inherently introduce noise. Convergence diagnostics (e.g., R^\hat{R}) showed values close to 1 for all parameters, indicating successful convergence of the chains. The model’s $R^2$ equals 0.967 indicates that 96.7% of the data variance is explained, demonstrating an excellent fit. The effective sample size (neffn_{\text{eff}}) for key parameters exceeded 500, ensuring the robustness of the inferences. Sensitivity analysis on priors revealed that minor adjustments to prior specifications did not significantly affect the model's predictions, demonstrating high model stability.

@@

For the variable `date`, we hypothesize that the prices of Loblaws "No Name" products fluctuate over time, reflecting seasonal trends, market dynamics, and promotional activities. This hypothesis is based on historical observations, where prices often vary due to promotions or stock-outs. Long-term trends, on the other hand, may reflect broader economic conditions or changes in production costs.
The variable size captures the impact of product quantity on pricing. We expect larger-sized products, such as "Extra Large," to generally have higher prices due to increased quantities. However, economies of scale may result in lower unit prices for larger sizes, while smaller sizes, despite lower overall costs, may have higher unit prices. This pricing strategy is common in retail to ensure accessibility for different consumer groups while optimizing profit margins.

@@

# Results

Our results are summarized in @tbl-model.

@@

```{r, fig.width=8, fig.height=3}
#| echo: false
#| eval: true
#| label: fig-results
#| fig-cap: Prediction of average price for Loblaws' No Name Grocery by type and by size
#| warning: false

@@

```{r, fig.width=8, fig.height=3}
#| echo: false
#| eval: true
#| label: fig-results2
#| fig-cap: Prediction of average price for Loblaws' No Name Grocery by type and by size
#| warning: false
# with `product_name`, `adjusted_price`, and credible intervals (lower_ci and upper_ci)
p7 <- ggplot(prediction_data, aes(x = size, y = adjusted_price)) +
  geom_boxplot(aes(lower = adjusted_lower_ci, upper = adjusted_upper_ci), color = "blue", fill = "lightblue", alpha = 0.5) +
  geom_hline(yintercept = mean(prediction_data$adjusted_price), linetype = "dashed", color = "black") +
  labs(
    title = "Prediction of Average Product Prices by Product Name",
    x = "Product Name",
    y = "Predicted Price ($)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 5),
    plot.title = element_text(size = 7, face = "bold", color = "#333333"),
    plot.subtitle = element_text(size = 6, color = "#333333"),
    legend.title = element_text(size = 5, color = "#333333"),
    legend.text = element_text(size = 5, color = "#333333"),
    legend.key.size = unit(0.5, "lines"),
    axis.title = element_text(size = 6, color = "#333333"),
    strip.text = element_text(size = 5)
)

p8 <- ggplot(prediction_data, aes(x = type, y = adjusted_price)) +
  geom_boxplot(aes(lower = adjusted_lower_ci, upper = adjusted_upper_ci), color = "blue", fill = "tomato", alpha = 0.5) +
  geom_hline(yintercept = mean(prediction_data$predicted_price), linetype = "dashed", color = "black") +
  labs(
    title = "Prediction of Average Product Prices by Product Name",
    x = "Product Name",
    y = "Predicted Price ($)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 5),
    plot.title = element_text(size = 7, face = "bold", color = "#333333"),
    plot.subtitle = element_text(size = 6, color = "#333333"),
    legend.title = element_text(size = 5, color = "#333333"),
    legend.text = element_text(size = 5, color = "#333333"),
    legend.key.size = unit(0.5, "lines"),
    axis.title = element_text(size = 6, color = "#333333"),
    strip.text = element_text(size = 5)
)

p7 + p8
```

@@

@@

Due to space limitations, we have chosen to display only the posterior and prior comparisons for a subset of variables in the posterior analysis. If we were to display all variables included in the model, the chart would become overly complex and difficult to interpret. Although the effects of multiple variables differ, focusing on this subset may give the impression that the model’s sensitivity to variations across different factors is limited. Thus, we have selected representative variables, including the intercept, size categories ("sizeLarge" and "sizeSmall"), and product types ("typeSnacks," "typeFlour/Rice," and "typeSeasoning"). The output results for other variables not shown are generally consistent with the trends observed here. Additionally, in regression models (such as stan_glmer), categorical variables (factors) are typically automatically converted into dummy variables. One category is set as the reference category (baseline), and its effect is included in the intercept, while the coefficients for other categories represent their offsets relative to the baseline. For example, in our model, if `Extra Large` is set as the baseline (default), it will not appear as a separate coefficient. The coefficients for `sizeLarge` and `sizeSmall` would then represent their deviations relative to `Extra Large`.

>>>>

---
title: "Does the Price Stability of 'No Name' Mean No Price Increases? A Strategy by Loblaws to Retain Customers"
subtitle: "Price Changes of Private-Label Brands: Stability, Strategy, and Consumer Impact"
author: 
  - Chenming Zhao
thanks: "Code and data are available at: [https://github.com/ChenmingZ2000/Loblaws-NoName-Grocery-Price-Trend-Analysis.git](https://github.com/ChenmingZ2000/Loblaws-NoName-Grocery-Price-Trend-Analysis.git).
        We gratefully acknowledge Jacob Filipp’s groceries dataset used in this analysis. The data is available at: [https://jacobfilipp.com/hammer/](https://jacobfilipp.com/hammer/)"
date: today
date-format: long
abstract: "This paper examines the pricing dynamics of Loblaws' private-label brand, No Name, in the context of rising grocery prices and consumer dissatisfaction in Canada. Using a hierarchical Bayesian model, we analyzed price fluctuations of No Name products across multiple categories to identify trends and potential strategies employed by Loblaws. Our findings shows that while prices exhibit minor fluctuations, No Name products have largely remained stable, suggesting a deliberate strategy by Loblaws to retain customers amidst growing competition and economic pressures. This study highlights the role of private-label brands in maintaining consumer loyalty and offers ideas about how large retailers balance affordability and profitability in an inflationary environment."
format: pdf
number-sections: true
bibliography: references.bib
pdf-engine: lualatex
---

@@

The Bank of Canada has noted that, since the economy began recovering from the pandemic, Canadians have experienced high inflation across many goods and services, with food prices rising faster than most other categories. Several factors have contributed to the surge in food costs, including supply chain disruptions, labor shortages, adverse weather conditions, and geopolitical events [see more details in @bankofcanada2024groceries and @statcan2022foodprices]. For example, in September 2022, annual food price inflation at grocery stores reached 11.4%, the fastest pace since 1981. In Toronto, food bank visits also set a new record, with 3.49 million visits recorded between April 2023 and April 2024, marking a significant increase from the previous year. This trend highlights the growing difficulty residents face in coping with the rising cost of living [see more details in @cbc2024foodbank].

Among Canada’s major grocery vendors, Loblaw Companies Limited has been at the center of widespread discussions and dissatisfaction due to price increases. As reported in @cbc2024people consumers have expressed frustration with the continued rise in food prices, particularly as inflation has begun to slow, raising questions about the pricing power of large grocery chains. Furthermore, Loblaw and its parent company, George Weston Limited, agreed to pay CAD 500 million to settle a class-action lawsuit related to bread price-fixing allegations, further exacerbating public dissatisfaction with the company's pricing strategies [see more details in @cbc2024].

The estimand of this study is the price fluctuations of grocery items under Loblaw’s private-label brand, No Name. As one of Canada’s largest grocery retailers, Loblaw’s No Name brand has a high level of market coverage and consumer recognition nationwide. Positioned as an affordable, high-value brand, No Name aligns closely with the needs of consumers facing inflation and rising food prices. It also serves as a key tool for Loblaw to increase customer retention. By studying No Name's pricing and market behavior, we aim to infer Loblaw’s overall pricing strategies.
Our analysis indicates that while most No Name products exhibit price fluctuations, their prices have remained relatively stable without significant increases. This suggests that Loblaw is likely maintaining control over No Name prices and using the PC Optimum loyalty program to prevent customer attrition.

The remainder of this paper is structured as follows. @sec-data introduces the data sources, preprocessing steps, variable selection criteria, and potential relationships between independent and dependent variables. @sec-model provides a concise explanation of the hierarchical Bayesian model, including the design rationale and the selection of priors and random effects, with more details on the model’s posterior in [Appendix -@sec-model-details]. @sec-result presents the main findings of the study, including a summary of the model output to support its accuracy and predictions of grocery price trends by different categories of “No Name” products through February 15, 2025. Finally, @sec-discussion includes the implications of the results, limitations of the data, complexities of incorporating real-world factors into the dataset, potential future pricing strategies by Loblaw, and steps for further research.

@@

To simulate, test, analyze, and clean the polling data, the statistical programming language R was employed [@citeR]. Key libraries that supported the data analysis include `tidyverse` [@tidyverse], `dplyr` [@dplyr], `arrow` [@arrow], `brms` [@brms_package] `knitr` [@knitr], `ggplot2` [@ggplot2], `kableExtra` [@kableExtra] and `patchwork` [@patch]. Additionally, the `rstanarm` [@rstanarm] library was utilized to implement multilevel modeling for prediction purposes. Also this analysis inherits the folder structure from Professor Rohan Alexander from University of Toronto [@rohan].

## Measurement
In the real world, pricing data for consumer goods is a dynamic reflection of market trends, regional preferences, and competitive strategies. The prices of products fluctuate based on supply chain disruptions, seasonal demand, and retail strategies, making it essential to track and analyze this information systematically. The Hammer dataset serves as a resource in understanding these trends, offering a detailed collection of product pricing data sourced from eight vendors. This dataset is embedded with complexities arising from the data collection process, competitive pricing strategies, and evolving consumer behavior.

Based on content on the Project Hammer website, the data collection process for @hammer is web scraping. This is one of the typical approaches used to construct structured datasets from online retail platforms. That is, retailers provide public access to their product inventories and prices through their websites, making them prime candidates for automated data extraction. Web scraping tools are deployed to extract product details such as names, brands, vendors, categories, unit prices, and quantities, while timestamps capture the exact moment of data retrieval to ensure time-sensitive accuracy. However, even with automated methods, the process is not without its challenges. Retailers often introduce barriers like CAPTCHA systems or dynamic content loading, which require solutions such as headless browsers or API integrations to ensure smooth data acquisition.

Sampling and representativeness are considerations in Hammer's data collection process.Inherent biases may exist. For instance, not all vendors update their online product catalogs in real time, which may introduce lag in capturing price changes or include the same data for different products. Once the raw data is collected, it undergoes cleaning and standardization. Product names, for instance, often include variations due to marketing strategies (e.g., "Club Size" or "Family Pack"), which are standardized to ensure consistency across entries. Similarly, data with missing values—such as prices or unit information—are flagged for further investigation.

@@

@fig-2 shows how average prices vary over time, segmented by product `type` (left) and `size` (right). In the left panel, we observe distinct pricing patterns across product types. For instance, "Seasoning" demonstrates higher volatility compared to "Flour/Rice," likely due to differences in demand elasticity. The "Milk" category shows occasional price dips, possibly influenced by the introduction of new products or promotional events.The right panel shows price trends segmented by size. "Extra Large" products generally have higher average prices but exhibit more unstable trends over time. In contrast, "Small" products show minimal price variation, possibly due to their inherently lower pricing or because their unit price is relatively higher compared to larger sizes.

@@

```{r, message=FALSE, warning=FALSE, fig.width= 7, fig.height= 3}
#| label: fig-2
#| fig-cap: Average Price Trend for Loblaws "No Name" Groceries Over Time Categorized in Type (left) and in Size (right).
#| echo: false
# Group data by date and product type, and calculate average price for each type per date
type_trend_data <- loblaws_data %>%
  group_by(date, type) %>%
  summarise(avg_price = mean(current_price, na.rm = TRUE)) %>%
  ungroup()

@@

The goal of our modeling strategy is to build a pricing model aimed at predicting changes in `current_price` over time while considering key variables that influence price fluctuations. The structure of the model combines fixed effects and random effects and includes multiple variables, such as `product_id`, product `size`, and product `type`. This modeling approach is designed to balance the complexity of the data while ensuring robust predictions. The hierarchical structure of the model allows us to capture variations at the level of individual products while identifying overall trends across different product sizes and types. It also enables us to forecast future price trends of Loblaws "No Name" groceries, observing whether the retailer's pricing behavior aligns with their commitments. Here, we provide a brief overview of the Bayesian analytical model used to investigate the price trends of Loblaws "No Name" products. Background details, model assumptions, and diagnostics are included in [Appendix -@sec-model-details].

@@

where $u_{\text{productID}[i]}$ is the random intercept for each product ID, capturing product-specific variations in pricing.

@@

The model was implemented using the `rstanarm` package and employed MCMC sampling for Bayesian inference. To validate the model, we conducted out-of-sample testing and calculated the predictive accuracy using RMSE (Root Mean Square Error), which was 0.83 on the log-transformed price scale. In the context of grocery price data, this level of error is acceptable, as market forces and external factors inherently introduce noise. Convergence diagnostics (e.g., $\hat{R}$) showed values close to 1 for all parameters, indicating successful convergence of the chains. The effective sample size (n_eff) for key parameters exceeded 500, ensuring the robustness of the inferences. Sensitivity analysis on priors revealed that minor adjustments to prior specifications did not significantly affect the model's predictions, demonstrating high model stability.

@@

For the variable `date`, we hypothesize that the prices of Loblaws "No Name" products fluctuate over time, reflecting seasonal trends, market dynamics, and promotional activities. This hypothesis is based on historical observations, where prices often vary due to promotions or stock-outs. Long-term trends, on the other hand, may reflect broader economic conditions or changes in production costs.

The variable size captures the impact of product quantity on pricing. We expect larger-sized products, such as "Extra Large," to generally have higher prices due to increased quantities. However, economies of scale may result in lower unit prices for larger sizes, while smaller sizes, despite lower overall costs, may have higher unit prices. This pricing strategy is common in retail to ensure accessibility for different consumer groups while optimizing profit margins.

@@

# Results {#sec-result}

## Model Output

Our model results are summarized in @tbl-model. This summary shows the effects of product attributes, including size and type, on price predictions. For instance, products categorized as "Large" are associated with a price reduction of 0.47 CAD compared to the baseline (Extra Large), while "Small" products show an even larger reduction of 0.98 CAD. Similarly, "Snacks" exhibit a negative effect of 0.39 CAD on price compared to the baseline type (Cooking Oil), while "Flour/Rice" shows a negligible effect. 
The model achieves an $R^2$ value of 0.967, indicating that it explains 96.7% of the variation in the data. This shows that the model effectively captures the main patterns in the data. The marginal $R^2$, which excludes the random effects, is 0.742, suggesting that fixed effects alone explain a substantial portion of the variance. The Intraclass Correlation Coefficient (ICC) is 0.1, indicating that only 10% of the variability is attributable to clustering by `product_id`, suggesting relatively low within-group correlation. Additional metrics support the model's robustness. The WAIC (6585.0) and LOOIC (6585.1) values confirm that the model balances predictive accuracy with generalizability, avoiding overfitting.

@@

## Prediction Output

Our analysis aims to predict the price trends of Loblaws' "No Name" products from November 20, 2024, to February 2025. The results of our model are illustrated in @fig-results1 and @fig-results2. @fig-results1 presents the predicted trends categorized by product type and size, providing a comparison across different products. The model predicts overall moderate price stability, with some variation between product types and sizes. @fig-results2 shows the prediction distribution categorized by product size.

At the product category level, the left panel of @fig-results1 shows that the price of Snacks remains relatively stable, while the other four categories exhibit comparatively larger fluctuations. These trends align with historical data. However, compared to historical data, the model underestimates the price volatility in certain categories. For instance, the historical data shows significant short-term price fluctuations in Seasoning, which are not fully reflected in the predicted results.

For product size, the right panel of @fig-results1 and @fig-results2 both indicate clear price stratification among Small, Large, and Extra Large products, with larger-sized products having higher average prices. The model successfully replicates a reasonable stratification structure, but the predicted price range is narrower, and the price volatility is lower compared to the historical observations.

```{r, fig.width=8, fig.height=3}
#| echo: false
#| eval: true
#| label: fig-results1
#| fig-cap: Prediction of average price trend for Loblaws' No Name Grocery by type and by size
#| warning: false

@@

```{r, fig.width=5, fig.height=3}
#| echo: false
#| eval: true
#| label: fig-results2
#| fig-cap: Prediction of average price distribution for Loblaws' No Name Grocery by size
#| warning: false

# with `product_name`, `adjusted_price`, and credible intervals (lower_ci and upper_ci)
ggplot(prediction_data, aes(x = size, y = adjusted_price)) +
  geom_boxplot(aes(lower = adjusted_lower_ci, upper = adjusted_upper_ci), color = "blue", fill = "tomato", alpha = 0.5) +
  geom_hline(yintercept = mean(prediction_data$adjusted_price), linetype = "dashed", color = "black") +
  labs(
    title = "Prediction of Average Product Prices by Product Name",
    x = "Product Name",
    y = "Predicted Price ($)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 5),
    plot.title = element_text(size = 7, face = "bold", color = "#333333"),
    plot.subtitle = element_text(size = 6, color = "#333333"),
    legend.title = element_text(size = 5, color = "#333333"),
    legend.text = element_text(size = 5, color = "#333333"),
    legend.key.size = unit(0.5, "lines"),
    axis.title = element_text(size = 6, color = "#333333"),
    strip.text = element_text(size = 5)
)

# Discussion{#sec-discussion}

## Price Forecast

In this paper, we focus on forecasting the price trends of Loblaws "No Name" products from November 20, 2024, to February 15, 2025. The analysis emphasizes the dynamics of prices categorized by product type and size, identifying behavioral patterns and differences between categories. We incorporated historical data like in @fig-1 and @fig-2 to ensure that the predictions align with observed past trends. Additionally, we designed the prediction dataset with constraints to simulate reasonable price changes—prices adjust at least every 14 days, with fluctuations ranging between 10% and 30% of the previous price. The results in @sec-result, as illustrated in @fig-results1, indicate that while the prices of Loblaws "No Name" products are expected to vary over time, the overall increase or decrease remains relatively limited. Both the predicted and historical data demonstrate similar trends across different categories. This suggests that retail pricing adjustments and predictions for similar product categories may follow patterns that are relatively consistent over time.

From the projected price fluctuations and distributions, the forecast suggests that there will be no significant price increases for Loblaws "No Name" products analyzed in this study from late November 2024 to mid-February 2025. This aligns with Loblaws' stated commitment to "maintain price stability" for No Name products whenever possible.

## Limitations of Existing Data

This section discusses the limitations of the data, data processing, and the proposed model, aiming to assess its robustness and the challenges in capturing the complexities of price changes. Each step, from data collection to model construction, involves certain assumptions and potential biases. Below is a detailed discussion of these aspects:

Firstly, the raw data used for observation and modeling is extensive and complex, with numerous duplicates and classification errors, as detailed in [Appendix -@sec-data-details]. One of the major challenges for the model is providing accurate predictions under limited accuracy and highly complex categorization. The complexity of the data makes it challenging for the model to provide clear interpretations. For example, the model might struggle to distinguish whether price fluctuations are caused by random events (such as promotions) or long-term trends (such as seasonal demand changes). This reduces the model's practical applicability, especially when decision-makers need straightforward conclusions.

The high variability in the data means that the model may struggle to accurately capture key trends and patterns. With a high level of noise, parameter estimates may be significantly affected, leading to wider prediction intervals and reduced reliability. For example, frequent fluctuations or outliers in the price data for certain products could cause the model's mean predictions to deviate from actual median trends. Misclassified product prices make it difficult for the model to correctly assess the impact of product characteristics on price. This may lower the model's ability to differentiate between categories, resulting in predictions that deviate from reality.

To reduce bias, we opted to clean duplicates and classification errors before modeling, attempting to standardize the data format. However, this greatly reduced the usable data volume. Unfortunately, since the true state of the data is unknown, the cleaned data may still contain inaccuracies. In the future, when faced with highly variable and complex categories, the model may need to incorporate additional random effects or interaction terms to account for data heterogeneity. However, such complexity may lead to overfitting, where the model fits the training data too closely but performs poorly on new data. This might also increase computational demands to impractical levels.

@loblaws2022response also emphasized the importance of regional and independent retailers in maintaining market competitiveness, particularly in non-urbanized areas. However, due to the dataset from @hammer only include the data from one community in Canada, our model cannot track the price behavior across regions, which may oversimplify the actual dynamics of the Canadian grocery market. If geographic information were incorporated, introducing regional effects or hierarchical modeling could enhance prediction accuracy, especially in capturing price trend differences between urban centers and less competitive markets.

## Real-World Impact

Another limitation of our analysis lies in its reliance on historical pricing data as the primary input. While this approach ensures consistency, it risks overlooking real-time market signals, such as sudden shifts in consumer preferences due to trends, unexpected supply chain disruptions for certain goods, or the introduction of new products with prices significantly different from similar items in the same category. This can result in predictions lagging behind actual market conditions, leading to inaccuracies in forecasting price volatility and trends and diminishing the model's real-time applicability. Additionally, the model does not account for external factors like inflation, competitor pricing, or macroeconomic indicators. These variables also influence retail prices, but due to a lack of data, they were excluded. Furthermore, the relatively narrow predicted price range suggests that the model may rely too heavily on prior assumptions, reducing its sensitivity to changes in price dynamics.

In future research, if sufficient data becomes available, we plan to enhance the model by incorporating dynamic factors or real-time updating mechanisms to respond quickly to market changes. This could include forecasting pricing for new products and simulating their potential impact on the market. However, like mentioned in @loblaws2022response price categorization in retail is often determined by consumer demand, packaging specifications, or branding strategies, which may not directly align with physical dimensions or averages of the products. This challenge is difficult to resolve, as we cannot predict changes in consumer preferences, meaning the results may not fully reflect actual market characteristics. The predictions will always lag behind real-world outcomes, with the goal being to move as close to reality as possible.

## Loblaws' Pricing Strategy

In @walrus2022grocer, the article highlights fairer supplier agreements, with one of the core goals of the code of conduct being to reduce conflicts and non-transparent practices between suppliers and retailers. This could provide greater market access for small and medium-sized suppliers, increasing the flexibility of private label supply chains, and would enable Loblaws to maintain price stability at lower supplier costs [see more details in @walrus2022grocer]. Combined with the low price volatility observed in our prediction results, this dynamic could further support Loblaws in offering competitive prices through its "No Name" brand, as these products primarily emphasize affordability. The code of conduct could help Loblaws maintain or improve No Name product quality without significantly increasing prices by reduced costs.

Based on the content from @canadiangrocer2023loyalty and @Li_Sexton_Xia_2006, we suggest that Loblaws positions its private label "No Name" at a lower price point to attract price-conscious consumers while gaining stronger bargaining power within the supply chain. This strategy likely allows Loblaws to stabilize or even reduce supplier costs while maintaining customer satisfaction with pricing. Furthermore, promotional activities such as discounts and loyalty rewards significantly influence consumer purchasing behavior and brand preferences. This aligns with Loblaws' approach to engaging consumers through its PC Optimum program.

Moreover, by using the extensive consumer data collected through PC Optimum, Loblaws can more accurately adjust product pricing and promotional strategies, continually refining its market positioning. Loyalty programs not only attract price-sensitive consumers but also increase the switching costs for customers moving to competitors, further boosting market share [see more details in @canadiangrocer2023loyalty]. Thus, the conjecture is, even if prices for other goods rise, Loblaws can mitigate negative consumer reactions by offering discounts, reward points, and stable pricing for its private label products. The "No Name" brand helps enhance consumer tolerance for price increases by maintaining an image of affordability. By preserving the perception of low prices, Loblaws reduces the visibility of price hikes, allowing it to sustain its market position and profitability without losing its overall customer base.

However, based on the report by @mcdonald-gibson_outraged_2024, Loblaws’ policies have not effectively appeased consumer sentiment. Consumers believe that Loblaws’ record profits amid rising food prices demonstrate corporate greed and disregard for consumer interests. Although the boycott has gained some attention, Loblaws’ stock price continues to rise, showing that the boycott has had limited financial impact on the company. When it comes to groceries, consumers have relatively limited choices. They usually choose stores that are convenient for their daily shopping needs. Furthermore, as a large corporation with multiple subsidiaries, such as Shoppers, it is challenging for consumers to completely boycott or significantly reduce their shopping frequency without affecting their daily lives. We believe that companies like Loblaws should soon make a balance between service, procurement costs, and pricing, minimizing unnecessary expenses and providing consumers with affordable groceries.

@@

# Data Collection Methodology Overview and Evaluation {#sec-data-details}

Observational data forms the core of our models; however, unlike experimental data, it is inherently limited by sampling biases, incomplete data, misclassification or duplicate classification of products, and inconsistent cross-product size standardization. Imagine you are assembling a complex machine, where each component must fit perfectly to deliver precision and efficiency. This is precisely how the Hammer Project operates—piecing together grocery pricing data to construct a unified picture of competition and pricing behaviors in Canada’s grocery market. In this section, we explore the foundation of Hammer Project’s data collection and analysis methodologies.

## Data Sources and Collection
The [Hammer Project](https://jacobfilipp.com/hammer/) collects its data through web scraping, capturing pricing information directly from the user interfaces of major grocery retailers’ websites without using APIs. Prices are specifically for the "in-store pickup" option and are based on a neighborhood in Toronto. The dataset includes eight vendors: Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart Canada, and Save-On-Foods.

However, relying on web scraping instead of APIs comes with limitations. APIs often provide more accurate and reliable historical price data, while web scraping can only capture data currently displayed on the website. For example, if a product is removed from the webpage after selling out rather than being marked as "out of stock," it may not be captured. Changes in website structure or server interruptions can also lead to inconsistencies, requiring researchers to perform additional data cleaning and supplementation. Furthermore, focusing solely on "in-store pickup" prices, without considering delivery or other purchasing methods, might fail to accurately reflect overall pricing strategies or market behaviors of grocers. This limitation narrows the dataset’s applicability. Additionally, collecting data from a single geographic area may restrict the generalizability of findings to other regions of Canada.

To ensure comprehensive coverage of grocery price trends, the dataset expanded its scope after the initial phase. Early data collection (from February 28 to July 10/11) focused on a small subset of products, but a broader variety of products was included starting July 10/11. This later expansion better reflects the diversity of products and pricing behaviors across vendors.
However, this sudden introduction of new data after July 10/11 poses challenges for analyzing price trends over time. It could create disruptions within categories, impacting the consistency of earlier patterns. Since the February to July period constitutes a significant portion of the current dataset (as of November 2024), researchers cannot exclude this period entirely to eliminate disturbances.

## Data Cleaning and Integration
@hammer processes its data into two core tables: a product table containing metadata and a raw table with time-series price data. These tables are joined using unique product identifiers. Challenges include managing duplicate entries for the same product and distinguishing between similar items. As noted on the Hammer Project website, a single product might appear multiple times under different categories, or subtle branding differences might complicate matching. Addressing these issues requires future data cleaning. However, since the data cannot be revisited retrospectively to correct inaccuracies from the time of collection, these challenges make ensuring complete data reliability more difficult.

Additionally, the Hammer Project dataset encourages users to explore targeted research questions. Examples include:

- `How do prices for common grocery bundles (e.g., eggs, milk, and bread) change over time and across vendors?`

- `Are "sales" genuine discounts, or do they follow price hikes?`

- `Which vendors consistently offer the lowest prices across comparable products?`

These questions align with the structured approach seen in polling methodology, where clear and purposeful design guides analysis. The transparency and flexibility of the dataset support academic and legal research, fostering a deeper understanding of Canada’s grocery industry.

@@

Due to space limitations, we have chosen to display only the posterior and prior comparisons for a subset of variables in the posterior analysis. If we were to display all variables included in the model, the chart would become overly complex and difficult to interpret. Although the effects of multiple variables differ, focusing on this subset may give the impression that the model’s sensitivity to variations across different factors is limited. Thus, we have selected representative variables, including the intercept, size categories (`sizeLarge` and `sizeSmall`), and product types (`typeSnacks`, `typeFlour/Rice`, and `typeSeasoning`). The output results for other variables not shown are generally consistent with the trends observed here. Additionally, in regression models (such as stan_glmer), categorical variables are typically automatically converted into dummy variables. One category is set as the reference category (baseline), and its effect is included in the intercept, while the coefficients for other categories represent their offsets relative to the baseline. For example, in our model, if `Extra Large` is set as the baseline (default), it will not appear as a separate coefficient. The coefficients for `sizeLarge` and `sizeSmall` would then represent their deviations relative to `Extra Large`.
